Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-17 SCP2008/paper/metamonads.tex
finish complete readthrough and spell check

\begin{abstract}
We use multi-stage programming, monads and Ocaml's
advanced module system to demonstrate how to eliminate all
abstraction overhead from generic programs while avoiding any inspection of
the resulting code.  We demonstrate this clearly with Gaussian Elimination as a 
representative family of symbolic and numeric algorithms. 
We parameterize our code to a great extent --
over domain, input and permutation matrix representations, 
determinant and rank tracking, 
pivoting policies, result types, etc. -- at no run-time cost.  Because
the resulting code is generated just right and not changed afterwards,
MetaOCaml guarantees that the generated code is well-typed.
We further demonstrate that various abstraction parameters (aspects)
can be made orthogonal and compositional, even in the presence of
name-generation for temporaries, and 
``interleaving'' of aspects.  We also show how to encode some
domain-specific knowledge so that ``clearly wrong'' compositions can
be rejected at or before generation time, rather than during
the compilation or running of the generated code.
\end{abstract}

@@

\subsection{MetaOCaml}

We build large code generators out of primitive generators using 
combinators. MetaOCaml, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the necessary
features: to construct a code expression, to combine them, and to
execute them. The following shows the simplest code generator |one|,
and the simplest code combinator\footnote{%
$\Longrightarrow$ under an expression shows the result of its evaluation}:

@@

We use MetaOCaml brackets |.<...>.| to generate code expressions,
i.e., to construct future-stage computations. MetaOCaml provides only
one mechanism of combining code expressions, by splicing one
piece of code into
another. The power of that operation, called escape |.~|, comes from
the fact that the expression to be spliced in (inlined) can be
computed: escape lets us perform an arbitrary immediate code-generating
computation \emph{while} we are
building a future-stage computation. The immediate computation in
|simplest_code| is the evaluation of the function |gen|, which in turn
applies |plus|. The function |gen| receives code expressions |.<x>.|
and |.<y>.| as arguments. At the generating stage, we can manipulate
code expressions as (opaque) values. The function |gen| returns a code
expression, which is inlined in the location of the escape. MetaOCaml can
print out code expressions, so we can examine the final generated code. It
has no traces of |gen| or |plus|: they are generation stage computations.

@@

By parametrizing our code, we can make the benefits of code generation
evident:

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields a function on |float|s, whereas
the latter expression is the code expression for a boolean function.
This clearly shows the separation of concerns for domain
operations.

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be a rather complex computation which we would rather not do
twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in an imperative context where the compiler
might not be able to determine whether lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}[commandchars=\\\{\}]
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization only saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not yet
have the |body|. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

Comparing the let-insertion in the generator
\begin{code}
let ce = (plus y one) in  plus ce (plus x ce)
\end{code}
with the corresponding code generating let-insertion for a future
stage
\begin{code}
letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
\end{code}
clearly shows the difference between  direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one, |let| became
``inverted''. For another, what used to be an expression that yields
a value, |init|, became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of parentheses around |fun| increases, the need to add and
then invoke the |k| continuation argument become increasingly annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

@@

let ret (a :'v) : ('p,'v) monad = fun s k -> k s a
let bind (m : ('p,'v) monad) (f : 'v -> ('p,'u) monad) : ('p,'u) monad
  = fun s k -> m s (fun s' b -> f b s' k)
let fetch s k = k s s  and  store v _ k = k v ()

@@

let ifL test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k -> 
  k s .< if .~test then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad}\label{ourmonad}
\end{figure}

@@

In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and a value, and
yields an answer of type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable for the environment
classifiers \cite{taha-environment} (such as the type variable |'c| 
in the type of |retN| in Figure~\ref{ourmonad}).
With type constructors taking more
and more arguments, it becomes increasingly difficult to read and write
types -- which we will be doing extensively when writing module
signatures in \S\ref{functors}. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these sorts of problems has been suggested by 
Jacques Garrigue on the Caml mailing list.
%% (cited from
%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
%% no date, no exact ref). 
%
We use a single type parameter |'p| to
represent all parameters of our monad (all parameters but the type of
the monadic value |'v|). The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type, as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

The operation |retN a| is the let-insertion operation, whose simpler
version we called |letgen| earlier. It is the first computation with a
control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Such behavior is symptomatic of control operators (in
particular, |abort|).  The name can be taken to mean \emph{return a
  Named computation}; the name allows for proper sharing, but
otherwise |retN| is used in writing generators in the same way as
|ret|. The type of |retN| is a specialization of the type of |ret|:
the computation |retN| deals specifically with code values. The types
of code values must include the environment classifier 
(such as |'c|), which denotes
the scope of free variables that may occur in the code value. The
argument type of |retN| and the answer type of |retN| computation must
have the same classifier -- which, informally, means that all free
variables in the argument of |retN| are preserved in the answer of
|retN|'s computation. When writing, for clarity, the type annotations
of |retN| we see the benefits of type-level records: we introduce a
new component of the record to specify the environment classifier
|'c|, and we update the |answer| component of the record to specialize
the answer type to be the type of code. The |state| component of the
record is not affected, and so remains hidden in the ellipsis |..|. The
number of parameters to the type constructor |monad| remains the same.

Finally, |runM| runs our monad, that is, given the initial state,
it performs the computation of
the monad and returns its result, which in our case is a code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|.
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
\end{code}
% param_code3 plus one;;
%
That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell (see App.~\ref{app:perform}).
\begin{code}
let param_code4 plus one =
  let gen x y = perform ce <-- retN (plus y one);
                        ret (plus ce (plus x ce))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
\end{code}
The function
|param_code4|, written using the |perform|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\subsection{Generating control statements}\label{ssec:control}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
can ``lift it'':
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
\end{code}
However we also need another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is apparent from the following example:
\begin{code}[commandchars=\\\{\}]
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
If we use |ifM'| to generate guarded array access code, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that code generation (such as the one in |retN|) is 
truly an effect, and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly.

@@

@@

Various aspects of our generator need to keep a state during
code generation (e.g., the name of the variable for the sign of
the determinant, \S\ref{sec:det}).

The simplest method of keeping the state of the generator is by using
mutable variables, private to each module (aspect). That would however
make our aspects stateful. Although we are generating imperative code,
we would like to keep our generators stateless and purely functional,
for ease of comprehension and reasoning.  Our main program may include
several generators referring to one particular aspect -- which may be
present in one shared instance or in several. That is of no concern if
the module is stateless, but with stateful modules, the issues of
aliasing or separate instantiation are a source of very subtle
problems.

We therefore chose a different way of maintaining generator state,
using a monad. We already need a monad for let-insertion; we now add state as
an additional monad parameter (Fig.~\ref{ourmonad}) and monadic
actions |fetch| and |store| to access that monadic state. The state is
threaded throughout the entire code-generation computation.

@@

The MLton team suggests a better approach: property 
lists~\cite{mlton-proplist}.  A property list also represents an
extensible object but by its dual, namely a list of fields.  The
initial object is just the empty list.  Unfortunately, we cannot apply
MLton's approach literally to our case. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e., fields of an extensible record. This
technique would make our aspects stateful modules, which we described
earlier as undesirable. MLton's approach to generating field
names also makes it difficult to store in the fields MetaOCaml's code
values, whose type contains a generic type variable, the environment
classifier. 
Fortunately, OCaml lets us build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g., |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols of
Lisp and Scheme.

Below is the implementation of our open records with manifest
naming: functions |orec_store| to add a new field to the open record
and |orec_find| to obtain the value associated with a particular field
name. Each ``field'' is characterized by a triple: an injection function, 
a projection function and the string name. The latter is used for 
printing error messages. For example, for the
determinant tracking aspect, this triple has the form
\begin{code}
let ip = 
  (fun x -> `TDet x), (function `TDet x -> Some x | _ -> None), "Det"
\end{code}
We combine these functions with monadic
actions to access monadic state and so obtain 
|mo_extend| and |mo_lookup| to store and retrieve one component 
of the monadic state.
\begin{code}
type ('a,'b) open_rec = ('a -> 'b) * ('b -> 'a option)  * string

let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec) )
   : 'b list -> 'a = 
   function [] -> raise Not_found
   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)

let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
   : 'b list =
  let () = 
    try let _ = lookup ip s in 
        failwith ("The field of an open record is already present: " ^ name)
    with Not_found -> () in
  (inj v)::s

@@

let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generation-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.  Our approach has the advantage of generating much
clearer error messages.

@@

For detailed reference, we present one particular Gaussian Elimination%
\footnote{For the purposes of this paper, we will use Gaussian Elimination (GE)
and LU decomposition (LU) as quasi-synonyms, even though we are well-aware
that GE can be used for other purposes, and LU can be performed by other
means than GE.}
algorithm, for an in-place LU-decomposition of an integer matrix with
full pivoting, returning the U-factor, determinant and rank. The
$n\times m$-matrix is represented as a flat vector, with 0-indexed
elements laid out in row-major format (C-style). The code below is
a hand-written implementation of the typical pseudo-code in
Numerical Analysis textbooks (see for example~\cite{BurdenFaires89}).

In OCaml, the matrix is represented by the value of the following type:
\begin{code}
type 'a container2dfromvector = {arr:('a array); n:int; m:int}

@@

let swap_cols a (n,m) c j =
  let end_vector = n * m in
  let rec loop col_c col_j =
    if col_j < end_vector then
      begin
        swap a col_c col_j;
        loop (col_c + m) (col_j + m)
      end
  in loop c j
\end{code}
We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
where |a| is the array component of the record |container2dfromvector|.
The above code defines three auxiliary functions, typically used in the
textbook pseudo-code: |swap a i j| swaps two elements; 
|swap_cols a (n,m) c j| swaps the column |c| with the column |j| in
the |(n,m)|-matrix |a|. The function 
|swap_rows| |a| |(n,m)| |(r,c) i| swaps the row |r| with the row |i| in 
the non-yet examined portion of the matrix, 
rectangular block |(r,c)-(n,m)|. We do not touch the 
elements to the left of the column |c| because they are all zeros. 
Since we know the layout of the matrix, we avoid 2D index
computations.

\begin{figure}
\begin{code}[numbers=left,numberblanklines=false,fontsize=\scriptsize]
let find_pivot a (n,m) r c =
  let pivot = ref None in               (* ((i,j), pivot_val) option *)
  begin
  for i = r to n-1 do
    for j = c to m-1 do
      let cur = a.(i*m+j) in
      if not (cur == 0) then
          match !pivot with
          | Some (_,oldpivot) ->
              if abs oldpivot > abs cur then
                 pivot := Some ((i,j), cur)
          | None -> pivot := Some ((i,j),cur)
  done; done;
  !pivot
  end

>>>>

\begin{abstract}
We use multi-stage programming, monads and Ocaml's
advanced module system to demonstrate how to eliminate all
abstraction overhead from generic programs while avoiding any inspection of
the resulting code.  We demonstrate this clearly with Gaussian Elimination as a 
representative family of symbolic and numeric algorithms. 
We parameterize our code to a great extent --
over domain, input and permutation matrix representations, 
determinant and rank tracking, 
pivoting policies, result types, etc. -- at no run-time cost.  Because
the resulting code is generated just right and not changed afterward,
MetaOCaml guarantees that the generated code is well-typed.
We further demonstrate that various abstraction parameters (aspects)
can be made orthogonal and compositional, even in the presence of
name-generation for temporaries, and 
``interleaving'' of aspects.  We also show how to encode some
domain-specific knowledge so that ``clearly wrong'' compositions can
be rejected at or before generation time, rather than during
the compilation or running of the generated code.
\end{abstract}

@@

\subsection{MetaOCaml and basic abstraction}\label{ssec:metaocaml}

We wish to build large code generators out of primitive generators using 
combinators. MetaOCaml, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the necessary
features: to construct a code expression, to combine them, and to
execute them. The following shows a simple code generator |one|,
and a simple code combinator\footnote{%
$\Longrightarrow$ under an expression shows the result of its evaluation}:

@@

We use MetaOCaml brackets |.<...>.| to generate code expressions,
i.e. to construct future-stage computations. MetaOCaml provides only
one mechanism for combining code expressions, by splicing one
piece of code into another. The power of that operation, called escape and
denoted |.~|, comes from
the fact that the expression to be spliced in (inlined) can be
computed: escape lets us perform an arbitrary \emph{code-generating}
computation \emph{while} we are
building a future-stage computation. The immediate computation in
|simplest_code| is the evaluation of the function |gen|, which in turn
applies |plus|. The function |gen| receives code expressions |.<x>.|
and |.<y>.| as arguments. At the generating stage, we can manipulate
code expressions as (opaque) values. The function |gen| returns a code
expression, which is inlined in the location of the escape. MetaOCaml 
conveniently can
print out code expressions, so we can examine the final generated code. It
has no traces of |gen| or |plus|: those are purely generation stage
computations.

@@

By parameterizing our code, we can make the benefits of code generation
evident:

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields a function on |float|s, whereas
the latter expression is a code expression for a boolean function.
This simple technique clearly shows how we can be polymorphic over
domain operations and yet still generate efficient domain-specific code,
thus achieving a proper separation of concerns.

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be a rather complex computation which we would rather not do
twice. We might be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in an imperative context where the compiler
might not be able to determine whether lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}[commandchars=\\\{\}]
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
The result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  This is because our |let|-insertion optimization only saved
the computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code, in a other words a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not yet
have the |body|. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

Comparing the let-insertion in the generator
\begin{code}
let ce = (plus y one) in  plus ce (plus x ce)
\end{code}
with the corresponding code generating let-insertion for a future
stage
\begin{code}
letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
\end{code}
clearly shows the difference between  direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one, |let| became
``inverted''. Secondly, what used to be an expression that yields
a value, |init|, became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of parentheses around |fun| increases considerably, the need to add and
then invoke the |k| continuation argument become increasingly annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that programmers of scientific software (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

@@

let ret (a :'v) : ('p,'v) monad = fun s k -> k s a
let bind (m : ('p,'v) monad) (f : 'v -> ('p,'u) monad) : ('p,'u) monad
  = fun s k -> m s (fun s' b -> f b s' k)

let fetch s k = k s s  and  store v _ k = k v ()

@@

let ifL test th el = ret .< if .~test then .~th else .~el >.
let ifM test th el = fun s k -> 
  k s .< if .~test then .~(th s k0) else .~(el s k0) >.
\end{code}
\caption{Our monad, helper functions and uses}\label{ourmonad}
\end{figure}

@@

In Figure~\ref{ourmonad}, the monad (yielding values of type |v|)
is implemented as a function of two
arguments: the state (of type |s|) and the continuation. The
continuation receives the current state and a value, and
yields an answer of type |w|.  The monad is polymorphic over the
three type parameters, which would require |monad| to be a type
constructor with three arguments. When we use this monad for code
generation, we will need yet another type variable for the environment
classifiers \cite{taha-environment} (such as the type variable |'c| 
in the type of |retN| in Figure~\ref{ourmonad}).
With type constructors taking more
and more arguments, it becomes increasingly difficult to read and write
types -- which we will be doing extensively when writing module
signatures in \S\ref{functors}. The fact that OCaml renames all type
variables when printing out types confuses matters further. An elegant
solution to these sorts of problems has been suggested by 
Jacques Garrigue on the Caml mailing list.
%% (cited from
%% \url{http://groups.google.com/group/fa.caml/msg/e80b1245702d6b24}
%% no date, no exact ref). 
%
We use a single type parameter |'p| to represent all parameters of our monad,
more precisely all parameters but the type of
the monadic value |'v|. The type variable |'p| is constrained to be
the type of an object with methods (fields) |state| and |answer|. The
object may include more fields, represented by |..|. Values of that
type are not part of our computations and need not exist. We merely
use the object type as an convenient way to specify extensible
\emph{type-level} records in OCaml.

@@

The operation |retN a| is the let-insertion operation, whose simpler
version we called |letgen| earlier. It is the first computation with a
control effect: indeed, the result of |retN a| is \emph{not} the
result of invoking its continuation |k|. Rather, its result is a |let|
code expression. Such behavior is symptomatic of control operators (in
particular, |abort|).  The name can be taken to mean \emph{return a
Named computation}; the name allows for proper sharing, but
otherwise |retN| is used in writing generators in the same way as
|ret|. The type of |retN| is a specialization of the type of |ret|:
the computation |retN| deals specifically with code values. The types
of code values must include the environment classifier 
(such as |'c|), which denotes
the scope of free variables that may occur in the code value. The
argument type of |retN| and the answer type of |retN| computation must
have the same classifier -- which, informally, means that all free
variables in the argument of |retN| are preserved in the answer of
|retN|'s computation. When writing, for clarity, the type annotations
of |retN| we see the benefits of type-level records: we introduce a
new component of the record to specify the environment classifier
|'c|, and we update the |answer| component of the record to specialize
the answer type to be a code type. The |state| component of the
record is not affected, and so remains hidden in the ellipsis |..|. The
number of parameters to the type constructor |monad| remains the same.

Finally, |runM| runs our monad, that is, given the initial state,
it performs the computation of
the monad and returns its result, which in our case is a code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|.
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
\end{code}
% param_code3 plus one;;
%
That may not seem like much of an improvement, but with the help of the
camlp4 pre-processor, we can introduce the |perform|-notation
\cite{metamonadsURL}, patterned after the |do|-notation of Haskell (see
App.~\ref{app:perform}).
\begin{code}
let param_code4 plus one =
  let gen x y = perform ce <-- retN (plus y one);
                        ret (plus ce (plus x ce))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.) ())>.
\end{code}
The function
|param_code4|, written using the |perform|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor converts the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\subsection{Generating control statements}\label{ssec:control}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
pieces of code rather than monadic values. We can ``lift it'':
\begin{code}
let ifM' test th el = perform
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc
\end{code}
However we also need another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is apparent from the following example:
\begin{code}[commandchars=\\\{\}]
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.) ())>.
\evalresult{.<fun a_1 i_2 ->}
\textcolor{red}{      if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
If we use |ifM'| to generate guarded array access code, the let-insertion
happens \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
after we have verified that the index is non-negative. This example
should make it clear that code generation (such as the one in |retN|) is 
truly an effect, and that we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly.

@@

One can think of this particular use of continuation as delimiting the 
``current scope'' of a block of code.  When constructing blocks
with a new scope, we use a fresh continuation; this allows us to generate
all named computations at the ``top'' of the current scope but no farther.

@@

Various aspects of our generator need to keep a state during
code generation (for example the name of the variable for the sign of
the determinant, \S\ref{sec:det}).
The simplest method of keeping such state is by using
mutable variables, private to each module (aspect). That would however
make our aspects stateful. Although we are generating imperative code,
we would like to keep our generators stateless and purely functional,
for ease of comprehension and reasoning.  Our main program may include
several generators referring to one particular aspect -- which may be
present in one shared instance or in several. That is of no concern if
the module is stateless, but with stateful modules, the issues of
aliasing or separate instantiation are a source of very subtle
problems.

We therefore chose a different way of maintaining generator state,
using a monad. We already saw that for let-insertions we could use
a continuation monad;  we now show how to use the state component
of our monad (Fig.~\ref{ourmonad}). The monadic actions |fetch| and
|store| are used to access that monadic state, which is
threaded throughout the entire code-generation computation.

@@

The MLton team suggests a better approach: property 
lists~\cite{mlton-proplist}.  A property list also represents an
extensible object but via its dual, namely a list of fields.  The
initial object is just the empty list.  Unfortunately, we cannot apply
MLton's approach literally to our case. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e. the fields of an extensible record. This
technique would make our aspects stateful modules, which we described
earlier as undesirable. MLton's approach to generating field names also makes
it difficult to store code values in those fields, as a code value has a type
which contains a generic type variable, the environment classifier. 
Fortunately, OCaml lets us build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g. |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols in Lisp and Scheme.

Below is the implementation of our open records with manifest
naming: functions |orec_store| to add a new field to an open record
and |orec_find| to obtain the value associated with a particular field
name. Each ``field'' is characterized by a triple: an injection function, 
a projection function and the string name. The latter is (only) used for 
printing error messages. For example, for the
determinant tracking aspect (\S\ref{sec:det}), this triple has the form
\begin{code}
let ip = 
  (fun x -> `TDet x), (function `TDet x -> Some x | _ -> None), "Det"
\end{code}
We combine these functions with monadic
actions to access monadic state and so obtain 
|mo_extend| and |mo_lookup| to store and retrieve one component 
of the monadic state.
\begin{code}
type ('a,'b) open_rec = ('a -> 'b) * ('b -> 'a option)  * string

let rec lookup ((_,prj,_) as ip:(('a,'b) open_rec)) : 'b list -> 'a = 
   function [] -> raise Not_found
   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)

let orec_store ((inj,_,name) as ip:(('a,'b) open_rec)) (v:'a) (s:'b list) 
   : 'b list =
  let () = 
    try let _ = lookup ip s in 
        failwith ("Field "^name^" of an open record is already present."
    with Not_found -> () in
  (inj v)::s

@@

let mo_lookup (ip:('a,'b) open_rec) : ('c, 'd) monad =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}
%
Currently, we check at generation-time that one should not add an
already existing field to an open record, nor should one attempt to
look up a field that does not exist. It is possible to make these
checks static.  Our approach has the advantage of generating \emph{much}
clearer error messages.

@@

For detailed reference, we present one particular Gaussian Elimination%
\footnote{For the purposes of this paper, we will use Gaussian Elimination (GE)
and LU decomposition (LU) as quasi-synonyms, even though we are well-aware
that GE can be used for other purposes, and LU can be performed by other
means than GE.}
algorithm, for an in-place LU decomposition of an integer matrix with
full pivoting, returning the U-factor, determinant and rank. The
$n\times m$-matrix is represented as a flat vector, with 0-indexed
elements laid out in row-major format (C-style). The code in 
Fig.~\ref{fig:GE-algorithm} is
a hand-written implementation of the typical pseudo-code in
Numerical Analysis textbooks (see for example~\cite{BurdenFaires89}).

In OCaml, the matrix is represented by a value of the following type:
\begin{code}
type 'a container2dfromvector = {arr:('a array); n:int; m:int}

@@

let swap_cols a (n,m) c j =
  let end_vector = n * m in
  let rec loop col_c col_j =
    if col_j < end_vector then
      begin
        swap a col_c col_j;
        loop (col_c + m) (col_j + m)
      end
  in loop c j
\end{code}
We refer to the $a_{ij}$-the element of the matrix as |a.(i*m+j)|,
where |a| is the array component of the record |container2dfromvector|.
The code above defines three auxiliary functions, typically used in 
textbook pseudo-code: |swap a i j| swaps two elements; 
|swap_cols a (n,m) c j| swaps column |c| with column |j| in
the |(n,m)|-matrix |a|. The function 
|swap_rows| |a| |(n,m)| |(r,c) i| swaps row |r| with row |i| in 
the non-yet examined portion of the matrix, 
rectangular block |(r,c)-(n,m)|. We do not touch the 
elements to the left of the column |c| because they are all zeros. 
Since we know the layout of the matrix, we avoid 2D index
computations.
